Binary companions of red giant stars with APOGEE: \\ catalog and companions that survive the common envelope

### 2018-02-06 paper/1-catalog/twoface-catalog.tex
update to aas 62
writing

\documentclass[modern, letterpaper]{aastex61}

@@

\newcommand{\nRC}{XX}
\newcommand{\nRGB}{XX}
\newcommand{\ntotal}{XX}
\newcommand{\ncompanions}{XX}

@@

All data used in this work comes from the publicly-available data release 14
(\DR) of the \apogee\ survey (\citealt{apogeedr14}), a component of the Sloan
Digital Sky Survey IV (\sdssiv; \citealt{sdss,Blanton:2017}).
\apogee\ is an infrared spectroscopic survey designed to map \RGB\ stars across
much of the Milky Way.
Targets are selected with simple color and brightness cuts, but the survey uses
fiber-plugged plates with a maximum of \todo{100?} fibers and a $\approx
1.5~{\rm deg}^2$ field of view, leading to ``pencil-beam''-like sampling of the
stellar distribution.
Data taken as part of the \apogee\ survey is reduced with the \todo{APOGEE
spectral parameter...} (\acronym{ASPCAP}; \citealt{Garcia-Perez:2016}), which
measures stellar parameters, chemical abundances, and radial velocities for each
target.
\todo{talk about visits, how the velocities measured...}

We use the primary data products from \apogee\ \DR\ (i.e. the \texttt{allStar}
and \texttt{allVisit} files) which contain 258475 unique source IDs
(\texttt{APOGEE\_ID}) and 1054381 unique visits.
We select all stars with $\geq 3$ visits that each pass a set of quality cuts,
described below.
For each visit, we require that the visit velocity uncertainty is $< 100~\kms$
(\texttt{VRELERR}) and the following bits are not set in the \texttt{STARFLAGS}
bitmask: \texttt{PERSIST\_HIGH}, \texttt{PERSIST\_JUMP\_POS},
\texttt{PERSIST\_JUMP\_NEG}, \texttt{VERY\_BRIGHT\_NEIGHBOR}, \texttt{LOW\_SNR}.
For each star, we require that $0 < \log g < 3.5$ and the following bits are not
set in the \texttt{ASPCAPFLAGS} bitmask: \texttt{STAR\_BAD}.
After these cuts, and the requirement of $\geq 3$ visits for a given star, we
are left with \todo{XX} visits for \todo{XX} unique stars (unique
\texttt{APOGEE\_ID}).
\todo{add numpy code for exact filter?}
\figurename~\ref{fig:nvisits} shows the number of stars in several bins of
number of visits that pass the above quality cuts: \todo{XX}\% of the stars in
\DR\ have $< 8$ visits.

% Notebook: "Numbers after cuts"
\begin{figure}[hp]
\begin{center}
\includegraphics[width=\textwidth]{nvisits.pdf}
\end{center}
\caption{%
TODO: \label{fig:nvisits}
}
\end{figure}

@@

\subsection{Orbit fitting and velocity modeling}\label{sec:fitting}

@@

We perform our individual system fits---that is, make our posterior
samplings---under the following assumptions:
\begin{description}
\item[no multiplets] All radial-velocity variations of the primary star are
  induced by a single companion.
  This is motivated by the idea that triple star systems are usually
  hierarchical so that the period of the inner binary is typically much shorter
  than the orbital period of the outer body (\todo{CITATION NEEDED}).
  At present, we ignore the possibility of higher-order multiple systems.
  \todo{revisit this - do we want to also sample over a long-term trend?}
\item[Kepler] Related to the first point, all velocity variations of the primary
  are gravitational, and we therefore ignore the possibility of intrinsic
  variation from, e.g., stellar oscillations.
\item[SB1] All spectra are single-lined; that is, we assume that the secondary
  is significantly fainter and is thus undetected in the spectra.
  This assumption is motivated by the fact that we expect \RGB\ stars to be
  substantially more luminous than their typical companion.
  However, there are known double-lined binary stars in the \apogee\ catalog
  (El-Badry et al., in prep.), and an unknown fraction of \RGB--\RGB\ binaries.
\item[simple noise model] Measurements are unbiased and noise estimates are
  correct up to an \todo{unknown} extra variance (the ``jitter'').
  All noise contributions result in Gaussian uncertainties on the individual
  radial-velocity measurements.
\end{description}

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in orbital parameters, given radial-velocity measurements
under the assumptions above.
Our parametrization is similar to the notation in \citet{Murray:2010}:
The radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, \phi_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,\phi_0,\omega,K,v_0)$---period, eccentricity, initial
phase, argument of pericenter, velocity semi-amplitude, Barycentric velocity---
and the true anomaly, $f$, is a function of the specified parameters (see
\sectionname~2 of \citealt{Price-Whelan:2017} or \eqname~63 in
\citealt{Murray:2010}).

\thejoker\ was designed for the extremely multi-modal pdfs expected when the
number of radial-velocity measurements of a source is small, or the data have
low signal-to-noise.
While other Markov Chain Monte Carlo (MCMC) methods have difficulty producing
independent samples with such data, \thejoker\ succeeds by brute force.
After generating an initial (very large) library of prior samples from an
assumed prior pdf (see below), the (typically multi-modal) likelihood is
evaluated at each sample and used to rejection sample.
In practice, given a number of requested samples for each star, the sampling
proceeds iteratively: since it is easier to accept samples when the data is
sparse or noisy, far more prior sample draws (and thus likelihood
evaluations) must occur under very constraining data.

@@

\subsection{Infer the jitter}

For each $n$ of $N$ RGB stars in APOGEE, we obtain $K$ posterior samples over
primary orbital parameters $\bs{w} = (P, e, \omega, \phi_0, K, v_0)$ and the
jitter parameter, $y = \ln s^2$, using \textit{The Joker}.
To obtain this sampling, we use an interim prior on the jitter parameter
parametrized by a mean and root-variance $\alpha_0 = (\mu_{y,0}, \sigma_{y,0})$.

@@

\begin{eqnarray}
    p(D_n \given \alpha) &=& \int \dd \theta_n \, p(D_n \given \theta_n) \,
      p(\theta_n \given \alpha)\\
    &=& \int \dd \theta_n \, p(D_n \given \theta_n) \, p(\theta_n \given \alpha) \,
      \frac{p(\theta_n \given D_n, \alpha_0)}{p(\theta_n \given D_n, \alpha_0)}\\
    &=& \int \dd \theta_n \, p(D_n \given \theta_n) \, p(\theta_n \given \alpha) \,
      p(\theta_n \given D_n, \alpha_0) \,
      \frac{p(D_n \given \alpha_0)}{p(D_n \given \theta_n) \,
      p(\theta_n \given \alpha_0)}\\
    &=& p(D_n \given \alpha_0) \, \int \dd \theta_n \,
      \frac{p(\theta_n \given \alpha)}{p(\theta_n \given \alpha_0)} \,
      p(\theta_n \given D_n, \alpha_0)\\
    &\approx& \mathcal{Z}_n \, \frac{1}{K_n} \,
      \sum_k^{K_n} \frac{p(\theta_{nk} \given \alpha)}{p(\theta_{nk} \given \alpha_0)}\\
    &=& \frac{\mathcal{Z}_n}{K_n} \,
      \sum_k^{K_n} \frac{p(y_{nk} \given \alpha)}{p(y_{nk} \given \alpha_0)}
\end{eqnarray}

@@

\subsection{Catalog / posterior samples whatever} \label{sec:full-catalog}

>>>>

\documentclass[modern, letterpaper]{aastex62}

@@

\newcommand{\nprior}{536,870,912}
\newcommand{\nposterior}{256}

@@

All data used in this work comes from the publicly-available data release 14
(\DR) of the \apogee\ survey (\citealt{apogeedr14}), a component of the Sloan
Digital Sky Survey IV (\sdssiv; \citealt{sdss,Blanton:2017}).
\apogee\ is designed to map stars across much of the Milky Way by obtaining
high-resolution ($R \sim 22500$) infrared ($H$-band) spectroscopy of primarily
\RGB\ stars.
Targets are selected with simple color and brightness cuts, but the survey uses
fiber-plugged plates with a maximum of 300 fibers per each $\approx 1.5~{\rm
deg}^2$ field of view, leading to ``pencil-beam''-like sampling of the stellar
distribution.
Data taken as part of the \apogee\ survey is reduced with the \apogee\ Stellar
Parameters and Chemical Abundances Pipeline (\acronym{ASPCAP};
\citealt{Garcia-Perez:2016}), which measures stellar parameters, chemical
abundances, and radial velocities for each target.
In order to meet signal-to-noise ratio requirements, most \apogee\ stars are
observed multiple times in a series of ``visits,'' typically with at least one
visit separated by a month or more in order to help identify binary stars.
\acronym{ASPCAP} iteratively solves for the relative velocities of each visit
spectrum while determining the optimal co-addition of the visit spectra.

We use the primary data products from \apogee\ \DR\ (i.e. the \texttt{allStar}
and \texttt{allVisit} files) which contain 258475 unique source IDs
(\texttt{APOGEE\_ID}) and 1054381 unique visits.
We select all stars with $\geq 3$ visits that each pass a set of quality cuts,
described below.
For each visit, we require that the visit velocity uncertainty is $< 100~\kms$
(\texttt{VRELERR}) and the following bits are not set in the \texttt{STARFLAGS}
bitmask: \texttt{PERSIST\_HIGH}, \texttt{PERSIST\_JUMP\_POS},
\texttt{PERSIST\_JUMP\_NEG}, \texttt{VERY\_BRIGHT\_NEIGHBOR}, \texttt{LOW\_SNR}.
For each star, we require that $0 < \log g < 3.5$ and the following bits are not
set in the \texttt{ASPCAPFLAGS} bitmask: \texttt{STAR\_BAD}.
After these cuts, and the requirement of $\geq 3$ visits for a given star, we
are left with \todo{XX} visits for \todo{XX} unique sources.
\figurename~\ref{fig:nvisits} shows the number of stars in several bins of
number of visits that pass the above quality cuts: \todo{XX}\% of the stars in
\DR\ have $< 8$ visits.

% Notebook: "Numbers after cuts"
\begin{figure}[hp]
\begin{center}
\includegraphics[width=\textwidth]{nvisits.pdf}
\end{center}
\caption{%
Number of \apogee\ stars in logarithmic bins of number of visits that pass the
quality cuts described in \sectionname~\ref{sec:data}.
In total, this work uses \todo{XX} visits and \todo{XX} unique sources.
\label{fig:nvisits}
}
\end{figure}

@@

\subsection{Orbit inference and velocity modeling}\label{sec:fitting}

@@

We perform our individual system fits---that is, make our posterior
samplings---under the following assumptions:
\begin{description}
\item[no multiplets] All radial-velocity variations of the primary star are
  induced by a single companion.
  This is motivated by the idea that triple star systems are usually
  hierarchical so that the period of the inner binary is typically much shorter
  than the orbital period of the outer body (\todo{CITATION NEEDED}).
  At present, we ignore the possibility of higher-order multiple systems.
  % \todo{revisit this - do we want to also sample over a long-term trend?}
\item[Kepler] Related to the first point, all velocity variations of the primary
  are gravitational, and we therefore ignore the possibility of coherent intrinsic variation from, e.g., stellar oscillations.
\item[SB1] All spectra are single-lined; that is, we assume that the secondary
  is significantly fainter and is thus undetected in the spectra.
  This assumption is motivated by the fact that we expect \RGB\ stars to be
  substantially more luminous than their typical companion.
  However, there are known double-lined binary stars in the \apogee\ catalog
  \citealt{El-Badry:2017}, and an expected but unknown fraction of \RGB--\RGB\
  binaries.
\item[simple noise model] Measurements are unbiased and noise estimates are
  correct up to an unknown extra variance.
  All noise contributions result in Gaussian uncertainties on the individual
  radial-velocity measurements.
  We often refer to the extra variance as a ``jitter'' parameter because for
  upper \RGB\ stars the inferred extra variance will be a combination of extra
  uncertainty and true astrophysical surface jitter (\todo{CITATION NEEDED}).
\end{description}

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in Keplerian orbital parameters, given radial-velocity
measurements under the assumptions listed above.
Our parametrization of the orbital elements is similar to the notation in
\citet{Murray:2010}:
The radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, M_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,M_0,\omega,K,v_0)$---period, eccentricity, mean
anomaly at a reference time $t_0$, argument of pericenter, velocity
semi-amplitude, Barycentric velocity---and the true anomaly, $f$, is a function
of time and the specified parameters (see \sectionname~2 of
\citealt{Price-Whelan:2017} or \eqname~63 in \citealt{Murray:2010}).

\thejoker\ was designed for the extremely multi-modal pdfs expected when the
number of radial-velocity measurements of a source is small, or the data are
sparse (in phase-coverage) or noisy.
While other Markov Chain Monte Carlo (MCMC) methods have difficulty producing
independent samples with such data, \thejoker\ succeeds by brute force:
After generating an initial (very large) library of prior samples from an
assumed prior pdf (see below), the (typically multi-modal) likelihood is
evaluated at each sample and used to rejection sample.
In practice, given a number of requested samples for each star, the sampling
proceeds iteratively: since it is easier to accept samples when the data is
sparse or noisy, far more prior sample draws (and thus likelihood evaluations)
must occur under very constraining data.

\subsection{Model selection}

@@

\subsection{Initial inidivual-system posterior samplings}
\label{sec:initsampling}

For all \todo{XX} \apogee\ stars with $\geq 3$ good visits (see
\sectionname~\ref{sec:data}), we use \thejoker\ to generate posterior samplings
for each star under the assumptions listed above (see
\sectionname~\ref{sec:fitting}).
We start by generating a library of \nprior\ prior samples generated under the
prior defined in \citet{Price-Whelan:2017}: uniform or isotropic in angle
parameters, uniform in log-period over the domain $[1, 32768]~{\rm day}$, and a
beta distribution over eccentricity with fixed parameters
(\citealt{Kipping:2013}).
For the prior over the extra variance or jitter parameter, $s$, we initially use
a Gaussian over the transformed parameter $y = \ln s^2$ with the mean and
standard deviation of the initial distribution set to $(\mu_{y,0}, \sigma_{y,0})
= (10.6, 3)$ in units of ${\rm m}~{\rm s}^{-1}$.

We request \nposterior\ posterior samples for each source, and generate the
samples using an iterative procedure that adapts to the \todo{informativeness}
of the data to more efficiently generate posterior samples.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples.
When this happens, we proceed in one of two ways:
\begin{enumerate}
    \item \emph{needs MCMC}: If just 1 posterior sample is returned after
    exhausting the full library of prior samples, or if multiple (but fewer than
    \nposterior) are returned that all lie within a single mode of the posterior
    pdf, the orbit-space that is consistent with the data is treated as
    effectively unimodal. In this case, we use the location of the returned
    sample, or the median over returned samples to generate a small ball of
    initial conditions and use the ensemble MCMC sampler (\citealt{Goodman:})
    implemented in \python\ (\citealt{Foreman-Mackey:}) to continue sampling
    until we obtain \nposterior\ samples.
    \item \emph{needs more prior samples}: If more than 1 posterior sample is
    returned after exhausting the full library of prior samples and the samples
    lie in multiple modes of the posterior pdf, the star is flagged. In this
    case, the only way forward is to generate more prior samples and continue
    running \thejoker\ for these sources to generate more posterior samples.
\end{description}

A full run of \thejoker\ on all \todo{XX} \apogee\ stars used in this work took
approximately 300 hours on a compute cluster with 448 cores, with the time
dominated by sources with many ($\gtrsim 10$) visits.

\subsection{Inferring the extra variance distribution}
\label{sec:inferjitter}

We use the per-source posterior samplings under the default prior described
above to infer the parameters of the assumed Gaussian distribution over the
log-jitter: $(\mu_y, \sigma_y)$.
This inference serves as a test-case for future work, where we intend to use the
independent posterior samplings to construct a hierarchical inference over
companion property distributions, and is also a test of the visit velocity
uncertainties reported in the \apogee\ data products.
If the catalog uncertainties are significantly underestimated, we expect the
inferred log-jitter distribution parameters to tend towards larger values.

To update the log-jitter distribution parameters, we maximize the marginal
likelihood of these population-level parameters.
We compute this marginal likelihood using the per-object posterior samples
re-weighted by the ratio of the value of the hyperprior evaluated at a new
location over the value of the default prior (see
\sectionname~\ref{sec:initsampling}).
This trick has been used in other hierarchical inferences as a way to
marginalize over the per-object parameters
(\citealt{Hogg:2008,Foreman-Mackey:...}), and we describe how to compute the
marginal likelihood in detail below.

\todo{Notation relative to above...}
For each $n$ of $N$ RGB stars in APOGEE, we obtain $K$ posterior samples over
primary orbital parameters $\bs{w} = (P, e, \omega, M_0, K, v_0)$ and the
jitter parameter, $y = \ln s^2$, using \thejoker.
To obtain this sampling, we use an interim (Gaussian) prior on the jitter
parameter parametrized by a mean and standard deviation $\alpha_0 = (\mu_{y,0},
\sigma_{y,0})$.

@@

\begin{eqnarray}
    p(D_n \given \alpha) &=& \int \dd \theta_n \, p(D_n \given \theta_n) \,
      p(\theta_n \given \alpha)\\
    &=& \int \dd \theta_n \, p(D_n \given \theta_n) \, p(\theta_n \given \alpha) \,
      \frac{p(\theta_n \given D_n, \alpha_0)}{p(\theta_n \given D_n, \alpha_0)}\\
    &=& p(D_n \given \alpha_0) \, \int \dd \theta_n \,
      \frac{p(\theta_n \given \alpha)}{p(\theta_n \given \alpha_0)} \,
      p(\theta_n \given D_n, \alpha_0)\\
    &\approx& \frac{\mathcal{Z}_n}{K_n} \,
      \sum_k^{K_n} \frac{p(y_{nk} \given \alpha)}{p(y_{nk} \given \alpha_0)}
\end{eqnarray}

Factor y from theta, so most priors cancel. Up to multiplicative constant all we need. Above just for one source, full population:

@@

TODO: apply to all stars with $>10$ visits

\subsection{Catalog of posterior samplings for APOGEE DR14 RGB stars}
\label{sec:fullcatalog}

TODO: re-do exactly \sectionname~\ref{sec:initsampling}, but with $(\mu_{y}, \sigma_{y}) = (9.5, 1.6)$ again in units of ${\rm m}~{\rm s}^{-1}$.

TODO: how many hit each category (flag) after the initial run
